Research Notes

### 2018-11-12 writeups/notes.tex
WIP

Let's restore the $\rho_1 g$ term now. For funsies, we begin by solving for
arbitrary stratification $\rho_0(z)$ first. The fluid equations to first order
reduce to
\begin{equation}
    \begin{split}
        \pd{\rho_1}{t} + \vec{u}_1 \cdot \p*{\vec{\nabla}\rho_0} &= 0,\\
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\vec{u}_1}{t} &= -\frac{\vec{\nabla}P_1}{\rho_0}
            - \frac{\rho_1 g}{\rho_0}
    \end{split}\label{eq:lin_incomp}
\end{equation}
We expect there to be some $z$ dependence in the amplitude, so we substitute
variables of form $e^{i(kx - \omega t)}$ and do not specify the $z$ dependence.
This gives us
\begin{equation}
    \begin{split}
        -i\omega \rho_1 - u_{1z}\pd{\rho_0}{z} &= 0,\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1 g}{\rho_g} &= 0.
    \end{split}
\end{equation}
We substitute $N^2 = -\frac{g}{\rho_0}\pd{\rho_0}{z}$ to obtain
\begin{subequations}
    \begin{align}
        -i\omega \rho_1 - u_{1z}\frac{\rho_0N^2}{g} &= 0,\label{eq:lin.cont}\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\label{eq:lin.div}\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\label{eq:lin.mom_x}\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1g}{\rho_0} &= 0.\label{eq:lin.mom_z}
    \end{align}
\end{subequations}
Eliminating $u_{1x}$ by substituting~\eqref{eq:lin.div}
into~\eqref{eq:lin.mom_x} and $\rho_1$ by substituting~\eqref{eq:lin.cont}
into~\eqref{eq:lin.mom_z} give
\begin{subequations}
    \begin{align}
        i\omega \pd{u_{1z}}{z} + \frac{k_x^2 P_1}{\rho_0} &= 0,
            \label{eq:lin.eq1}\\
        \p*{\omega^2 - N^2 }u_{1z} + \frac{i\omega}{\rho_0}\pd{P_1}{z} &= 0.
            \label{eq:lin.eq2}
    \end{align}
\end{subequations}
Finally, we multiply~\eqref{eq:lin.eq1} with $\rho_0$ and differentiate
$\mathrm{d}z$ and combine with~\eqref{eq:lin.eq2} to give
\begin{equation}
    \rtd{u_{1z}}{z} + \frac{1}{\rho_0}\pd{\rho_0}{z}\pd{u_{1z}}{z}
        + k_x^2\p*{\frac{N^2}{\omega^2} - 1}u_{1z} = 0.\label{eq:lin.incomp.gen}
\end{equation}

Let's now pick stratification $\rho \propto e^{-z/H}$
\autoref{eq:lin.incomp.gen} clearly has exponential solutions $e^{\kappa z}$ for
\begin{equation}
    \kappa^2 - \frac{\kappa}{H} + k_x^2\p*{\frac{N^2}{\omega^2} - 1} = 0.
\end{equation}
We permit complex $\kappa = \frac{1}{2H} + ik_z$, and from the above clearly
\begin{align}
    k_z^2 &= -\frac{1}{4H^2} + k_x^2\p*{\frac{N^2}{\omega^2} - 1},\nonumber\\
    \omega^2 &= \frac{N^2k_x^2}{k_x^2 + k_z^2 +
        \frac{1}{4H^2}}.\label{eq:strat_disp}
\end{align}
Thus the eigenfunctions are
\begin{equation}
    \begin{split}
        u_{1z} &= e^{z/2H} e^{i(k_zz + k_xx - \omega t)},\\
        u_{1x} &= -\frac{k_z + i/2H}{k_x} u_{1z},\\
        \rho_1 &= \frac{i \rho_0}{H\omega} u_{1z},\\
        P_1 &= -\frac{\rho_0 \omega}{k_x^2}\p*{k_z + i/2H}u_{1z}.
    \end{split}\label{eq:strat_eigens}
\end{equation}

@@

From here, it's a simple matter of taking limits of existing anelastic
approximations. We will start with the full system from Glatzmaier's book:
\begin{align}
    0 &= \vec{\nabla} \cdot \p*{\rho_0 \vec{u}_1},\\
    \rho_1 &= \frac{\rho_0}{\gamma P_0}P_1 - \frac{\rho_0}{c_p}S_1
        = \at{\pd{\rho_0}{P_0}}_S P_1 + \at{\pd{\rho_0}{S_0}}_P S_1,\\
    N^2 &= \frac{g}{c_p}\rd{S_0}{z},\\
    0 &= \pd{\vec{u}_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\vec{u}_1
        + \vec{\nabla}\p*{\frac{P_1}{\rho_0} + \Phi_1}
        + \at{\pd{\rho_0}{S_0}}P \frac{g}{\rho_0}S\hat{z}
        - \nu \p*{\nabla^2 \vec{u}_1 + \frac{1}{3}\vec{\nabla}\p*{\vec{\nabla}
            \cdot \vec{u}_1}},\\
    0 &= \pd{S_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}S_1
        + u_{1z}\rd{S_0}{z}
        - \frac{1}{\rho_0 T_0} \vec{\nabla} \cdot \p*{\kappa \rho_0 T_0
            \vec{\nabla} S_1}.
\end{align}
We obviously take $\Phi_1 = 0$ the Cowling approximation. Of course, in our
problem, we may drop all the viscous terms from the momentum equation and energy
equation, which immediately leaves us with
\begin{align}
    \vec{\nabla} \cdot \p*{\rho_0 \vec{u}_1} &= 0,\\
    \pd{\vec{u}_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\vec{u}_1
        + \vec{\nabla} \p*{\frac{P_1}{\rho_0}} - \frac{g}{c_p}S_1 \hat{z}
            &= 0,\\
    \pd{S_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}S_1
        &= 0.
\end{align}
Note that when accounting for $N^2 = \frac{g}{c_p}\rd{S_0}{z}$ so $c_p = S_0$
immediately gives us Sutherland's equations from his 2011 paper.

We now make one further approximation: we have already taken the low Mach number
approximation, so we will approximate $\rho_1 = \frac{\rho_0}{\gamma P_0}P_1 -
\frac{\rho_0}{S_0}S_1 \approx -\frac{\rho_0}{S_0}S_1$, which allows us to
eliminate $S$. We must be careful in taking this approximation (and indeed, it
may be unjustified), since the difference term $\vec{\nabla}\frac{P_1}{\rho_0} -
\frac{\vec{\nabla}P_1}{\rho_0} = -\frac{P_1 \vec{\nabla}\rho_0}{\rho_0^2}$ is
now small as well. What to do with it is dictated by the requirement that
horizontal momentum be conserved. With this in mind, the momentum equation
must be
\begin{equation}
    \pd{\vec{u}_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\vec{u}_1
        + \frac{\vec{\nabla}P_1}{\rho_0} + \frac{\rho_1}{\rho_0}g\hat{z} = 0.
\end{equation}
We may verify that horizontal momentum is conserved by multiplying through with
$\rho_0$ again.

@@

\section{The Anelastic/Boussinesq Approximations (old)}

@@

@@

We now have three equations for four variables, $\vec{u}_1, \rho_1, P_1$. We
must introduce a fourth equation, a thermodynamic equation. For an adiabatic
process $P\rho^{-\gamma} \propto P^{1-\gamma}T^\gamma$ is constant. We thus
introduce the concept of the \emph{potential temperature}
\begin{equation}
    \theta = T\p*{\frac{P_0}{P}}^\kappa.
        \label{eq:pot_temp}
\end{equation}
For an adiabatic process, $\rd{\theta}{t} = 0$. Motivated by this, we use
\begin{equation}
    \begin{split}
        \pd{1}{\rho_0}\pd{\rho_0}{z} &= \frac{1}{\gamma P_0}\pd{P_0}{z}
            - \frac{1}{\theta_0}\pd{\theta_0}{z},\\
        \frac{\rho_1}{\rho_0} &= \frac{1}{\gamma} \frac{P_1}{P_0}
            - \frac{\theta_1}{\theta_0},
    \end{split}
\end{equation}
to give the momentum equation form
\begin{equation}
    \rd{\vec{u}_1}{t} = -\vec{\nabla}\p*{\frac{P_1}{\rho_0}}
        + \frac{P_1}{\rho_0}\p*{\frac{1}{\theta_0}\vec{\nabla}\theta_0}
        + \vec{g} \frac{\theta_1}{\theta_0}.\label{eq:mom_unsimp}
\end{equation}

@@

The anelastic fluid equations thus read
\begin{equation}
    \begin{split}
        \vec{\nabla} \cdot \p*{\rho_0\vec{u}} &= 0,\\
        \pd{\vec{u}_1}{t} + \vec{\nabla}\p*{\frac{P_1}{\rho_0}}
            - \vec{g}\frac{\theta_1}{\theta_0} &= 0,\\
        \pd{\theta_1}{t} + \p*{\vec{u} \cdot \vec{\nabla}}\theta_0 &= 0.
    \end{split}\label{eq:lin_anelastic}
\end{equation}

>>>>

Let's restore the $\rho_1 g$ term now. For funsies, we begin by solving for
arbitrary stratification $\rho_0(z)$ first. The fluid equations to first order
reduce to
\begin{equation}
    \begin{split}
        \pd{\rho_1}{t} + \vec{u}_1 \cdot \p*{\vec{\nabla}\rho_0} &= 0,\\
        \vec{\nabla} \cdot \vec{u}_1 &= 0,\\
        \pd{\vec{u}_1}{t} &= -\frac{\vec{\nabla}P_1}{\rho_0}
            - \frac{\rho_1 g}{\rho_0}
    \end{split}\label{eq:lin_incomp}
\end{equation}
We expect there to be some $z$ dependence in the amplitude, so we substitute
variables of form $e^{i(kx - \omega t)}$ and do not specify the $z$ dependence.
This gives us
\begin{equation}
    \begin{split}
        -i\omega \rho_1 - u_{1z}\pd{\rho_0}{z} &= 0,\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1 g}{\rho_g} &= 0.
    \end{split}
\end{equation}
We substitute $N^2 = -\frac{g}{\rho_0}\pd{\rho_0}{z}$ to obtain
\begin{subequations}
    \begin{align}
        -i\omega \rho_1 - u_{1z}\frac{\rho_0N^2}{g} &= 0,\label{eq:lin.cont}\\
        iku_{1x} + \pd{u_{1z}}{z} &= 0,\label{eq:lin.div}\\
        -iw u_{1x} + \frac{ik_x P_1}{\rho_0} &= 0,\label{eq:lin.mom_x}\\
        -iw u_{1z} + \frac{1}{\rho_0}\pd{P_1}{z} +
            \frac{\rho_1g}{\rho_0} &= 0.\label{eq:lin.mom_z}
    \end{align}
\end{subequations}
Eliminating $u_{1x}$ by substituting~\eqref{eq:lin.div}
into~\eqref{eq:lin.mom_x} and $\rho_1$ by substituting~\eqref{eq:lin.cont}
into~\eqref{eq:lin.mom_z} give
\begin{subequations}
    \begin{align}
        i\omega \pd{u_{1z}}{z} + \frac{k_x^2 P_1}{\rho_0} &= 0,
            \label{eq:lin.eq1}\\
        \p*{\omega^2 - N^2 }u_{1z} + \frac{i\omega}{\rho_0}\pd{P_1}{z} &= 0.
            \label{eq:lin.eq2}
    \end{align}
\end{subequations}
Finally, we multiply~\eqref{eq:lin.eq1} with $\rho_0$ and differentiate
$\mathrm{d}z$ and combine with~\eqref{eq:lin.eq2} to give
\begin{equation}
    \ptd{u_{1z}}{z} + \frac{1}{\rho_0}\pd{\rho_0}{z}\pd{u_{1z}}{z}
        + k_x^2\p*{\frac{N^2}{\omega^2} - 1}u_{1z} = 0.\label{eq:lin.incomp.gen}
\end{equation}

Let's now pick stratification $\rho_0 \propto e^{-z/H}$
\autoref{eq:lin.incomp.gen} clearly has exponential solutions $e^{\kappa z}$ for
\begin{equation}
    \kappa^2 - \frac{\kappa}{H} + k_x^2\p*{\frac{N^2}{\omega^2} - 1} = 0.
\end{equation}
We permit complex $\kappa = \frac{1}{2H} + ik_z$, and from the above clearly
\begin{align}
    k_z^2 &= -\frac{1}{4H^2} + k_x^2\p*{\frac{N^2}{\omega^2} - 1},\nonumber\\
    \omega^2 &= \frac{N^2k_x^2}{k_x^2 + k_z^2 +
        \frac{1}{4H^2}}.\label{eq:strat_disp}
\end{align}
Thus the eigenfunctions are
\begin{equation}
    \begin{split}
        u_{1z} &= e^{z/2H} e^{i(k_zz + k_xx - \omega t)},\\
        u_{1x} &= -\frac{k_z + i/2H}{k_x} u_{1z},\\
        \rho_1 &= \frac{i \rho_0}{H\omega} u_{1z},\\
        P_1 &= -\frac{\rho_0 \omega}{k_x^2}\p*{k_z + i/2H}u_{1z}.
    \end{split}\label{eq:strat_eigens}
\end{equation}

@@

From here, it's a simple matter of taking limits of existing anelastic
approximations. We will start with the full system from Glatzmaier's book:
\begin{align}
    0 &= \vec{\nabla} \cdot \p*{\rho_0 \vec{u}_1},\\
    \rho_1 &= \frac{\rho_0}{\gamma P_0}P_1 - \frac{\rho_0}{c_p}S_1
        = \at{\pd{\rho_0}{P_0}}_S P_1 + \at{\pd{\rho_0}{S_0}}_P S_1,\\
    N^2 &= \frac{g}{c_p}\rd{S_0}{z},\\
    0 &= \pd{\vec{u}_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\vec{u}_1
        + \vec{\nabla}\p*{\frac{P_1}{\rho_0} + \Phi_1}
        + \at{\pd{\rho_0}{S_0}}P \frac{g}{\rho_0}S\hat{z}
        - \nu \p*{\nabla^2 \vec{u}_1 + \frac{1}{3}\vec{\nabla}\p*{\vec{\nabla}
            \cdot \vec{u}_1}},\\
    0 &= \pd{S_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}S_1
        + u_{1z}\rd{S_0}{z}
        - \frac{1}{\rho_0 T_0} \vec{\nabla} \cdot \p*{\kappa \rho_0 T_0
            \vec{\nabla} S_1}.
\end{align}
We obviously take $\Phi_1 = 0$ the Cowling approximation. Of course, in our
problem, we may drop all the viscous terms from the momentum equation and energy
equation, which immediately leaves us with
\begin{align}
    \vec{\nabla} \cdot \p*{\rho_0 \vec{u}_1} &= 0,\\
    \pd{\vec{u}_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\vec{u}_1
        + \vec{\nabla} \p*{\frac{P_1}{\rho_0}} - \frac{g}{c_p}S_1 \hat{z}
            &= 0,\\
    \pd{S_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}(S_0 + S_1)
        &= 0.
\end{align}
Note that when accounting for $N^2 = \frac{g}{c_p}\rd{S_0}{z}$ so $c_p = S_0$
immediately gives us Sutherland's equations from his 2011 paper.

We now make one further approximation: we have already taken the low Mach number
approximation, so we will approximate $\rho_1 = \frac{\rho_0}{\gamma P_0}P_1 -
\frac{\rho_0}{S_0}S_1 \approx -\frac{\rho_0}{S_0}S_1$, which allows us to
eliminate $S$. With this in mind, the momentum equation can be rewritten
\begin{equation}
    \pd{\vec{u}_1}{t} + \p*{\vec{u}_1 \cdot \vec{\nabla}}\vec{u}_1
        + \vec{\nabla}\p*{\frac{P_1}{\rho_0}}
        + \frac{\rho_1}{\rho_0}g\hat{z} = 0.
\end{equation}
We may verify that horizontal momentum is conserved by multiplying through with
$\rho_0$ again.

@@

\section{The Anelastic/Boussinesq Approximations}

@@

This follows a step by step derivation of the usual anelastic approximation that
is used in atmospheric sciences. It's not the most careful treatment but should
give some intuition as to how the equation of state is handled.

@@

We now have three equations for four variables, $\vec{u}_1, \rho_1, P_1$. We
must introduce a fourth equation, a thermodynamic equation. For an adiabatic
process $P\rho^{-\gamma} \propto P^{1-\gamma}T^\gamma$ is constant. We thus
introduce the concept of the \emph{potential temperature} (just the entropy)
\begin{equation}
    \theta = T\p*{\frac{P_0}{P}}^\kappa.
        \label{eq:pot_temp}
\end{equation}
For an adiabatic process, $\rd{\theta}{t} = 0$. Motivated by this, we use
\begin{equation}
    \begin{split}
        \frac{1}{\rho_0}\pd{\rho_0}{z} &= \frac{1}{\gamma P_0}\pd{P_0}{z}
            - \frac{1}{\theta_0}\pd{\theta_0}{z},\\
        \frac{\rho_1}{\rho_0} &= \frac{1}{\gamma} \frac{P_1}{P_0}
            - \frac{\theta_1}{\theta_0},
    \end{split}
\end{equation}
to give the momentum equation form
\begin{equation}
    \rd{\vec{u}_1}{t} = -\vec{\nabla}\p*{\frac{P_1}{\rho_0}}
        + \frac{P_1}{\rho_0}\p*{\frac{1}{\theta_0}\vec{\nabla}\theta_0}
        + \vec{g} \frac{\theta_1}{\theta_0}.\label{eq:mom_unsimp}
\end{equation}

@@

The anelastic fluid equations thus read
\begin{equation}
    \begin{split}
        \vec{\nabla} \cdot \p*{\rho_0\vec{u}_1} &= 0,\\
        \pd{\vec{u}_1}{t} + \vec{\nabla}\p*{\frac{P_1}{\rho_0}}
            - \vec{g}\frac{\theta_1}{\theta_0} &= 0,\\
        \pd{\theta_1}{t} + \p*{\vec{u} \cdot \vec{\nabla}}\theta_0 &= 0.
    \end{split}\label{eq:lin_anelastic}
\end{equation}
